---
title: "Sports Analytics Project"
author: "Shashwat Mishra - sdm4329"
date: "2026-04-17"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


---
title: "WNBA 2028 Player Projection Model V2"
author: "Your Team Names"
date: "April 2026"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Libraries

```{r libraries}
library(dplyr)
library(tidyr)
library(ggplot2)
library(broom)
library(knitr)
library(kableExtra)
```

## Step 1: Load and Clean Merged Dataset

```{r load_data}
master <- read.csv('/Users/shashwatmishra/Downloads/all_years_merged.csv') %>%
  rename(season      = Season,
         player_name = Name,
         team        = TeamAbbreviation) %>%
  mutate(
    ppg    = Points    / GamesPlayed,
    rpg    = Rebounds  / GamesPlayed,
    apg    = Assists   / GamesPlayed,
    bpg    = Blocks    / GamesPlayed,
    spg    = Steals    / GamesPlayed,
    tpg    = Turnovers / GamesPlayed,
    min_pg = Minutes   / GamesPlayed
  ) %>%
  filter(Minutes >= 100)

cat("Total rows:", nrow(master), "\n")
cat("Seasons:", paste(sort(unique(master$season)), collapse = ", "), "\n")
cat("Unique players:", n_distinct(master$player_name), "\n")
```

## Step 2: Add Age

```{r add_age}
players_2025 <- read.csv('/Users/shashwatmishra/Downloads/WNBA Players 2025 - Sheet1.csv') %>%
  select(player_name = PLAYER, age_2025 = AGE)

master <- master %>%
  left_join(players_2025, by = "player_name") %>%
  mutate(age = age_2025 - (2025 - season)) %>%
  filter(!is.na(age))

cat("Rows after age join:", nrow(master), "\n")
cat("Players retained:", n_distinct(master$player_name), "\n")
```

## Step 3: Create Lagged Features

```{r lagged_features}
model_data <- master %>%
  arrange(player_name, season) %>%
  group_by(player_name) %>%
  mutate(
    lag1_ortg = lag(OnOffRtg, 1),
    lag1_drtg = lag(OnDefRtg, 1),
    lag1_ppg  = lag(ppg,      1),
    lag1_rpg  = lag(rpg,      1),
    lag1_apg  = lag(apg,      1),
    lag1_ts   = lag(TsPct,    1),
    lag1_usg  = lag(Usage,    1),
    lag1_efg  = lag(EfgPct,   1),
    lag1_tpg  = lag(tpg,      1),
    lag1_spg  = lag(spg,      1),
    lag1_bpg  = lag(bpg,      1)
  ) %>%
  ungroup() %>%
  filter(!is.na(lag1_ortg))

cat("Model dataset rows:", nrow(model_data), "\n")
cat("Seasons covered:", paste(sort(unique(model_data$season)), collapse = ", "), "\n")
cat("Unique players:", n_distinct(model_data$player_name), "\n")
```

## Step 4: Train/Test Split (80/20)

```{r train_test}
set.seed(42)
train_idx <- sample(nrow(model_data), 0.8 * nrow(model_data))
train_set <- model_data[ train_idx, ]
test_set  <- model_data[-train_idx, ]

cat("Training rows:", nrow(train_set), "\n")
cat("Test rows:", nrow(test_set), "\n")
```

## Step 5: ORTG and DRTG Models

```{r models}

library(car)
ortg_model <- lm(OnOffRtg ~ lag1_ortg + lag1_ppg + lag1_ts +
                              lag1_usg + lag1_apg + lag1_tpg + age,
                 data = train_set)

drtg_model <- lm(OnDefRtg ~ lag1_drtg + lag1_rpg + lag1_spg +
                              lag1_bpg + lag1_tpg + age,
                 data = train_set)

summary(ortg_model)
summary(drtg_model)

vif(ortg_model)
vif(drtg_model)
```

## Step 5b: Trimmed Models (Significant Predictors Only)

```{r trimmed_models}
ortg_model_v2 <- lm(OnOffRtg ~ lag1_ortg + lag1_ppg + lag1_apg,
                    data = train_set)

drtg_model_v2 <- lm(OnDefRtg ~ lag1_drtg + lag1_spg + lag1_bpg,
                    data = train_set)

summary(ortg_model_v2)
summary(drtg_model_v2)
```

## Step 5c: Final Model With Efficiency Check

```{r final_models}
ortg_model_v3 <- lm(OnOffRtg ~ lag1_ortg + lag1_ppg + 
                                lag1_apg + lag1_ts,
                    data = train_set)

drtg_model_v3 <- lm(OnDefRtg ~ lag1_drtg + lag1_spg + lag1_bpg,
                    data = train_set)

summary(ortg_model_v3)
summary(drtg_model_v3)
```

## Step 6: Age Analysis

```{r age_analysis}
library(ggplot2)

# ORTG vs age
ggplot(master, aes(x = age, y = OnOffRtg)) +
  geom_point(alpha = 0.3, color = "#e05c3a") +
  geom_smooth(method = "lm", color = "#e05c3a", se = TRUE) +
  geom_smooth(method = "loess", color = "black", linetype = "dashed", se = FALSE) +
  labs(title = "Offensive Rating vs Age",
       subtitle = "Linear fit (red) vs LOESS curve (dashed)",
       x = "Age", y = "OnOffRtg") +
  theme_minimal()

# DRTG vs age
ggplot(master, aes(x = age, y = OnDefRtg)) +
  geom_point(alpha = 0.3, color = "#3a7de0") +
  geom_smooth(method = "lm", color = "#3a7de0", se = TRUE) +
  geom_smooth(method = "loess", color = "black", linetype = "dashed", se = FALSE) +
  labs(title = "Defensive Rating vs Age",
       subtitle = "Linear fit (blue) vs LOESS curve (dashed)",
       x = "Age", y = "OnDefRtg") +
  theme_minimal()

# Also check average rating by age bucket
master %>%
  mutate(age_bucket = cut(age, breaks = c(20, 24, 27, 30, 33, 36, 45),
                          labels = c("21-24", "25-27", "28-30", 
                                     "31-33", "34-36", "37+"))) %>%
  group_by(age_bucket) %>%
  summarise(
    n            = n(),
    avg_ortg     = round(mean(OnOffRtg, na.rm = TRUE), 1),
    avg_drtg     = round(mean(OnDefRtg, na.rm = TRUE), 1),
    avg_netrtg   = round(mean(OnOffRtg - OnDefRtg, na.rm = TRUE), 1)
  ) %>%
  print()
```

## Step 7: Test Set Evaluation

```{r evaluation_final}
test_set$pred_ortg <- predict(ortg_model_v3, newdata = test_set)
test_set$pred_drtg <- predict(drtg_model_v3, newdata = test_set)

cat("ORTG Train MSE:", round(mean(ortg_model_v3$residuals^2), 4), "\n")
cat("ORTG Test MSE: ", round(mean((test_set$OnOffRtg - test_set$pred_ortg)^2, na.rm=TRUE), 4), "\n")
cat("DRTG Train MSE:", round(mean(drtg_model_v3$residuals^2), 4), "\n")
cat("DRTG Test MSE: ", round(mean((test_set$OnDefRtg - test_set$pred_drtg)^2, na.rm=TRUE), 4), "\n")
```

## Step 8: Project 2028

```{r projections_final}
proj_2025 <- master %>%
  filter(season == 2025, Minutes >= 300) %>%
  group_by(player_name) %>%
  slice_max(order_by = Minutes, n = 1) %>%
  ungroup() %>%
  rename(
    lag1_ortg = OnOffRtg,
    lag1_drtg = OnDefRtg,
    lag1_ppg  = ppg,
    lag1_apg  = apg,
    lag1_ts   = TsPct,
    lag1_spg  = spg,
    lag1_bpg  = bpg
  ) %>%
  mutate(age = age + 3)

proj_2025$proj_ortg_2028 <- predict(ortg_model_v3, newdata = proj_2025)
proj_2025$proj_drtg_2028 <- predict(drtg_model_v3, newdata = proj_2025)

final_v3 <- proj_2025 %>%
  mutate(proj_netrtg_2028 = proj_ortg_2028 - proj_drtg_2028) %>%
  select(player_name, team, age,
         proj_ortg_2028, proj_drtg_2028, proj_netrtg_2028) %>%
  arrange(desc(proj_netrtg_2028)) %>%
  mutate(Rank = row_number())

print(final_v3, n = 30)
```

## Step 9: Final Rankings Table

```{r table_final}
final_v3 %>%
  select(Rank,
         Player  = player_name,
         Team    = team,
         Age     = age,
         `Proj ORTG` = proj_ortg_2028,
         `Proj DRTG` = proj_drtg_2028,
         `Net RTG`   = proj_netrtg_2028) %>%
  mutate(across(where(is.numeric), ~ round(., 1))) %>%
  kbl(align = c("c","l","c","c","c","c","c"),
      caption = "2028 WNBA Player Projections — Ranked by Net Rating") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = TRUE,
                font_size = 10) %>%
  row_spec(0, bold = TRUE)
```

## Step 10: Export Files

```{r exports_final}
write.csv(train_set,
          '/Users/shashwatmishra/Downloads/WNBA_train_set_v2.csv',
          row.names = FALSE)

write.csv(test_set,
          '/Users/shashwatmishra/Downloads/WNBA_test_set_v2.csv',
          row.names = FALSE)

write.csv(final_v3 %>%
            select(Rank, player_name, team, age,
                   proj_ortg_2028, proj_drtg_2028, proj_netrtg_2028),
          '/Users/shashwatmishra/Downloads/WNBA_2028_Projections_Final_V3.csv',
          row.names = FALSE)

cat("Exported train set:", nrow(train_set), "rows\n")
cat("Exported test set:", nrow(test_set), "rows\n")
cat("Exported projections:", nrow(final_v3), "players\n")
```

